---
title: "Houston Heat - Heat Related Illness Prevention"
subtitle: "April 15, 2026"
author: Andrea Reed & Kimberly Williams
format: html
---

```{r}
#| output: FALSE
library(tidyverse)
library(sf)        # shapefile --> a set of files that work together
library(sfdep)     # Chloropleth Map and Moran's I
library(spmodel)   # spautor model
library(viridis)   # colorblind friendly colors
library(patchwork) # put graphs together
```

You have been contracted by the Houston Health Department as a junior data analyst. The city is increasingly concerned about the impact of heat on public health (it's Houston after all) — especially among vulnerable populations. Your team’s mission is to analyze public health data related to **heat related deaths** in Houston and identify the key contributing factors. Your findings will help shape public awareness campaigns and guide local policy decisions.

## Data

```{r}
broken_dat <- st_read("Data/HoustonHeat.shp")  # sf object
dat <- st_make_valid(broken_dat)
```

Variables:

- Geometry - Coordinate info for the corresponding Houston census block groups
- Count - The number of heat-related illnesses reported last summer
- Population - The population of the census block
- NOAC - % of census block with no air conditioning 
- MED_AGE - Median age of people in census block 
- HispanicPC - % of Hispanic individuals in block 
- BlackPCT - % of Black individuals in block 
- under5PCT - % of individuals under age of 5 in block 
- over65PCT - % of individuals over age 65 in block 
- povertyPCT - % of individuals in poverty in block 
- alonePCT - % of individuals in block who live alone 
- MinTemp - Average daily minimum temperature last summer

Your objective is to analyze the data to identify contributing factors to heat-related illnesses as well as visualize high-risk areas across the greater Houston area. Your deliverable for this project is a 1-page (8.5x11 sheet) public-facing health flyer for a general audience, with insights or safety tips drawn from your analysis. Your flyer must contain the following:

- A model-based map of high and low risk areas across the city (not just the raw data)
- A list of factors that (significantly) contribute to increased or decreased risk (note that there are racial factors in here and this is a public-facing flyer so be sensitive and avoid harmful stereotypes)
- Suggestions on how to stay safe

If you don’t know what a flyer might look like, go to images.google.com and search something like “public health flyer for heat related illness” and you’ll get the idea. Canva is a great place to start with templates. Upload a PDF of flyer to gradescope.

```{r}
# option 1: filter out blocks without counts or population
RRdat1 <- dat |>
  mutate(RR = Count / (Population*(sum(Count)/sum(Population)))) |>
  mutate(log_RR = log(RR)) |>
  filter(is.finite(log_RR)) # filter out infinitives or NA
# option 2: filter out places without population, 
#           and add small constant to account for areas without any heat incidents
RRdat2 <- dat |>
  filter(Population > 0) |> # Removes 0 population to avoid division by zero
  mutate(RR = (Count + 0.1) / (Population * (sum(Count) / sum(Population)))) |>
  mutate(log_RR = log(RR))

RRdat <- RRdat2 # determine pick here!!
```


```{r}
ggplot(data=RRdat, mapping=aes(fill=log_RR)) +
  geom_sf(data = dat, fill = "lightgrey", color = "white") + # Background (all tracts)
  scale_fill_viridis() +
  theme_minimal() +
  geom_sf() +
  labs(title = "Chloropleth Map of log(Relative Risk)",
       fill = "log(RR)")
```

```{r}
model_formula <- RR ~ NOAC + MED_AGE + HispanicPC + BlackPCT + 
                 under5PCT + over65PCT + povertyPCT + alonePCT + MinTemp
car_model <- spautor(model_formula,
                     data = RRdat,
                     spcov_type="car")

resids = resid(car_model)
decorr_resid = rstandard(car_model)
fitted = fitted.values(car_model)

RRdat_fr <- RRdat |>
  mutate(resids = resids, decorr_resid = decorr_resid, fitted = fitted)
```

# LINE assumptions from CAR model

### Linearity

```{r}
rrlm <- lm(model_formula, data = RRdat)
car::avPlots(rrlm)
```


### Independence

```{r}
moran_test <- RRdat %>%
  st_contiguity() %>% #Extract the neighbors of each observation
  global_moran_test(x=decorr_resid, #Variable we are looking at
                    nb=.,
                    wt=st_weights(.))
moran_test
```
Nice it's independent

### Normality

```{r}
ks <- ks.test(decorr_resid, "pnorm", mean=0, sd=1)
pval <- formatC(ks$p.value, digits = 4)

hr <- ggplot(data.frame(resids = decorr_resid), aes(x = resids)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "orchid",
                 alpha = 0.7, color = "orchid") +
  stat_function(fun = dnorm, args = list(mean = mean(decorr_resid), 
                                         sd = sd(decorr_resid)),
                linewidth = 0.7, color = "darkred", linetype = "dashed") +
  labs(title = paste0("Histogram of Decorrelated Residuals", 
                      "\nKS test p-value: ", pval),
       x = "Decorrelated Residuals",
       y = "Density") +
  theme_minimal()
hr
```
Lol not normal at all

### Equal Variance

```{r}
bp_test <- lmtest::bptest(
  decorr_resid ~ log_RR,
  data = RRdat
)
bp_pval <- formatC(bp_test$p.value, digits = 4)

fr <- ggplot(data.frame(fitted = fitted, resid = decorr_resid),
       aes(x = fitted, y = decorr_resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Fitted Values",
       y = "Residuals",
       title = paste0("Fitted vs Decorrelated Residuals\n",
                    "BP Test p-value: ", bp_pval)) +
  theme_minimal()
fr
```
Hahaha also super heteroskedastic. Look at that trumpet!

```{r}
pseudoR2(car_model)
```

A pseudo-$R^2$ value of 0.04868105 means that roughly 5% of the variability in log(Relative Risk) can be explained by the car model.

```{r}
summary(car_model)
```


